Local Type Inference for Polarised System F with Existentials

### 2024-05-07 oopsla24/tex/01-introduction.tex
Edit: sections 1-5

In this paper, we extend the local type inference algorithm to a language with
both universal and existential quantifiers, which can both be instantiated
impredicatively. This combination of features broke a number of the invariants
which traditional type inference algorithms depend on, and required us to invent
new algorithms which combine both unification and (surprisingly)
the dual---anti-unification.

@@

\paragraph{Type Inference Algorithm (\cref{sec:algorithm})}
    We give a local type inference algorithm
    which supports both first-class existential and universal quantifiers, both
    of which can be instantiated impredicatively. To evade the undecidability
    results surrounding type inference for \systemf, we introduce \fexists---a
    variant of \CBPV \cite{levy2006:cbpv}, which lets us formulate a subtyping
    relation which is still decidable.

\paragraph{Declarative Specification (\cref{sec:declarative-system,sec:proofs})}
    We give a declarative type system (again, based on \CBPV) to
    serve as a specification of our algorithm, and we prove our algorithm is
    sound and complete with respect to the declarative type system. The
    specification makes it easy to see that all type applications (for
    $\forall$-elimination) and all packs (for $\exists$-introdution) are
    inferred.

\paragraph{Employment of the Anti-unification (\cref{sec:antiunification})}
    The introduction of impredicative existentials breaks some of the
    fundamental invariants of HM-style type inference. As a result, it needs to
    mix unification with \emph{anti-unification}---a dual operation which has
    been used in the literature on term rewriting systems, but which has never
    appeared in the context of polymorphic type inference.

\paragraph{Extrapolation of the Inference Framework to Other Type Systems
          (\cref{sec:extensions})}
    We explore the design space further to show how the
    same type inference scheme can be applied to
    work with different type systems. In particular,
    \begin{enumerate*}
      \item[(i)] in combination with \emph{bidirectional} typechecking---the 
        approach used in the original local type inference paper of 
        \citet{pierce2000:local} that minimizes the number of needed annotations;
      \item[(ii)] in systems with different subtyping rules, such as  
        the one used in \citet{zhao22:elementary}---a system that
        permits explicit type applications;
      \item[(iii)] in systems with \emph{bounded} polymorphic quantifiers.
        That is, one can only instantiate a quantifier with a 
        type that is a subtype of a given bound.
    \end{enumerate*}

\ilyam{Is the following needed since we have the contributions? I just wanted to introduce the appendix somewhere...}
\paragraph{\textbf{Structure of this Paper}}
In \cref{sec:overview}, we discuss the language of the \fexists system, providing
examples, outlining limitations, and summarizing the main ideas behind the
algorithm. \Cref{sec:declarative-system} presents a formal definition of the
declarative specification and its key properties. In \cref{sec:algorithm}, the
type inference algorithm is introduced, detailing a hierarchy of subroutines;
upon perusing this section, one could \emph{implement} the type inference
algorithm. \Cref{sec:proofs} overviews the algorithm's correctness proof,
explaining its structure and how the declarative system's features make it
decidable. \Cref{sec:extensions} covers potential \fexists extensions and future
work directions. Finally, \cref{appendix} contains comprehensive formal
definitions and proofs of lemmas and theorems used in the paper; in instances
where the full definitions might not be of immediate interest, we direct the
reader to the appendix.

>>>>

@@

\paragraph{Declarative Type System Specification (\cref{sec:declarative-system,sec:proofs})} 
    We devise \fexists---a second-order polymorphic type system that supports first-class
    existential and universal quantifiers, both of which can be instantiated
    impredicatively. To evade the undecidability results surrounding subtyping
    for \systemf, we base the system on \CBPV \cite{levy2006:cbpv}: we
    use \emph{polarization} of types and terms to formulate the restrictions
    that make typing decidable.

    We give a \emph{declarative} specification of local type inference in this
    system (\ie the typing information is only available at a limited scope)
    based on \emph{subtyping} relation. The specification makes it easy to see
    what is and what is not inferrable in the system.

    To show that our restrictions do not unduly limit the expressiveness of the
    language, we prove that every \fexists term can be embedded into \systemf
    and vice versa.

\paragraph{Type Inference Algorithm (\cref{sec:algorithm})} We give a type
  inference algorithm implementing the declarative specification. Because our
  system supports both existential and universal quantifiers, our implementation
  of the subtyping relation requires the use of \emph{anti-unification}---a dual
  to unification that finds the most specific generalization rather than the
  most general unifier. To our knowledge, anit-unification has never been used
  in the context of polymorphic type inference.

  The type inference algorithm itself is a purely local one, in the style of
  \citet{pierce2000:local}. The locality can be seen from the fact that 
  none of the unification variables introduced as part of type inference
  escape the scope of a single function application.

\paragraph{Theoretical results: correctness of the algorithm (\cref{sec:proofs})}
  We prove that our algorithm is terminating and both sound and complete with
  respect to the declarative specification. This proof of soundness and completeness
  of the algorithm required us to introduce an intermediate system in-between
  the declarative and the algorithmic systems, and furthermore, the
  soundness and completeness proofs are mutually recursive with each other.

% \paragraph{Extrapolation of the inference framework to other type systems (\cref{sec:extensions})}
%     We explore the design space further to show how the
%     same type inference scheme can be applied to
%     work with different type systems. In particular,
%     \begin{enumerate*}
%       \item[(i)] in combination with \emph{bidirectional} typechecking---the 
%         approach used in the original local type inference paper of 
%         \citet{pierce2000:local} that minimizes the number of needed annotations;
%       \item[(ii)] in systems with different subtyping rules, such as  
%         the one used in \citet{zhao22:elementary}---a system that
%         permits explicit type applications;
%       \item[(iii)] in systems with \emph{bounded} polymorphic quantifiers.
%         That is, one can only instantiate a quantifier with a 
%         type that is a subtype of a given bound.
%     \end{enumerate*}

% \ilyam{Is the following needed since we have the contributions? I just wanted to introduce the appendix somewhere...}
\paragraph{\textbf{Structure of this Paper}}
In \cref{sec:overview}, we discuss the language of the \fexists system, providing
examples, outlining limitations, and summarizing the main ideas behind the
algorithm. \Cref{sec:declarative-system} presents a formal definition of the
declarative specification and its key properties. In \cref{sec:algorithm}, the
type inference algorithm is introduced, detailing a hierarchy of subroutines;
upon perusing this section, one could \emph{implement} the type inference
algorithm. \Cref{sec:proofs} overviews the algorithm's correctness proof,
explaining its structure and how the declarative system's features make it
decidable. \Cref{sec:extensions} covers potential \fexists extensions and future
work directions. Finally, \cref{appendix} contains comprehensive formal
definitions and proofs of lemmas and theorems used in the paper; in instances
where the full definitions might not be of immediate interest, we direct the
reader to the appendix.

### 2024-05-07 oopsla24/tex/03-declarative-system.tex
Edit: sections 1-5

@@

  The latter rule is for the \emph{annotated} binder,
  and thus, the type of the bound $[[x]]$ is given, 
  however, the rule must check that this type is a
  a supertype of the inferred type of the application. 
  This check is done by invoking the subtyping judgment
  $[[Γ ⊢ ↑iQ ≤ ↑iP]]$.
  This judgment is more restrictive than checking bare 
  $[[Γ ⊢ iP ≥ iQ]]$, however, it is necessary
  to make the algorithm complete as it allows us to preserve
  certain invariants throughout the algorithm to ensure
  the decidability of the produced unification task.
  In \cref{sec:proof-subtyping}, we discuss these invariants in
  more detail, and in \cref{sec:weakening-invariant} we suggest 
  a way to relax them.

>>>>

\paragraph{Let binders}
  Rule \ruleref{\ottdruleDTVarLetLabel} is a standard rule for typing let
  binders: we infer the type of the bound value and continue the typing of the
  computation in the extended context.

  Rule \ruleref{\ottdruleDTCVarLetLabel} associates a variable with a computation.
  The inferred type for this computation ($[[iM]]$) must be castable to a
  shifted positive type $[[↑iP]]$, with $[[iP]]$ then assigned to the bound
  variable $[[x]]$ to type the let binder's body. Like all annotated
  constructors, we also verify the annotation type $[[iP]]$'s well-formedness.

@@

  The latter rule is for the \emph{annotated} binder,
  and thus, the type of the bound $[[x]]$ is given, 
  however, the rule must check that this type is a
  a supertype of the inferred type of the application. 
  This check is done by invoking the subtyping judgment
  $[[Γ ⊢ iM' ≤ ↑iP]]$.
  % This judgment is more restrictive than checking bare 
  % $[[Γ ⊢ iP ≥ iQ]]$, however, it is necessary
  % to make the algorithm complete as it allows us to preserve
  % certain invariants throughout the algorithm to ensure
  % the decidability of the produced unification task.
  % In \cref{sec:proof-subtyping}, we discuss these invariants in
  % more detail, and in \cref{sec:weakening-invariant} we suggest 
  % a way to relax them.
